% ============================================================
% Section 3: Epistemic System
% ============================================================

\section{Epistemic System}
\label{sec:epistemic}

The \texttt{EpistemicHead} is the central component that differentiates
\aletheion{} from conventional LLMs. It receives the final hidden states
from the transformer and produces the \textbf{epistemic tomography} ---
a structure with 30+ fields capturing the complete epistemic state of
each token.

\subsection{Epistemic Tomography}

\begin{definition}[Epistemic Tomography]
\label{def:tomography}
For each token $x_t$ in a sequence of batch size $B$ and length $T$,
the tomography $\mathcal{T}_t$ is a tuple:

\begin{equation}
    \mathcal{T}_t = \big(q_1, q_2, \bm{c}, \kappa, \phi, \bm{v}, \tau, \bm{e}, \gamma, \bm{s}, \ldots\big)
\end{equation}

where $q_1 \in [0,1]$ is the aleatoric uncertainty, $q_2 \in [0,1]$ the
epistemic uncertainty, $\bm{c} \in [0,1]^5$ the manifold 5D coordinates,
$\kappa \in [0,1]$ the MAD confidence, $\phi \in \R$ the manifold health,
$\bm{v} \in \R^5$ the intentionality vector, $\tau > 0$ the adaptive
temperature, and the remaining fields are optional (tiers 1--3).
\end{definition}

\subsection{Uncertainty Decomposition}

Following \cite{der2009aleatory}, we decompose total uncertainty into two
orthogonal components:

\begin{align}
    q_1 &= \sigma\big(\text{MLP}_{q_1}(\bh)\big) \in [0, 1] & \text{(aleatoric uncertainty)} \\
    q_2 &= \sigma\big(\text{MLP}_{q_2}(\bh)\big) \in [0, 1] & \text{(epistemic uncertainty)}
\end{align}

\begin{itemize}
    \item \textbf{$q_1$ (aleatoric)}: Captures the inherent ambiguity of
    the context --- ``The next token could be several equally valid
    things.'' Irreducible by more data.

    \item \textbf{$q_2$ (epistemic)}: Captures the model's lack of
    knowledge --- ``I don't know what the next token is.'' Reducible
    with more training.
\end{itemize}

The $q_1$ and $q_2$ gates are implemented as compact MLPs with
configurable depth (default 2 layers, hidden dimension 64):

\begin{equation}
    \text{QGate}(\bh) = \sigma\big(\bm{W}_2 \cdot \text{ReLU}(\bm{W}_1 \bh + \bm{b}_1) + \bm{b}_2\big)
\end{equation}

with $\bm{W}_1 \in \R^{64 \times d}$ and $\bm{W}_2 \in \R^{1 \times 64}$.

\subsection{Adaptive Temperature}

The sampling temperature is adaptively controlled by the combined
uncertainty signal $c = \text{clamp}(q_1 \cdot q_2, \epsilon, 1)$:

\begin{equation}
    \tau = \begin{cases}
        \tau_{\text{base}} / c & \text{if } c < \tau_{\text{threshold}} \\
        \tau_{\text{base}} & \text{otherwise}
    \end{cases}
\end{equation}

where $\tau_{\text{base}} = 1.0$ and $\tau_{\text{threshold}} = 0.5$.
When both uncertainties are high (product exceeds threshold), temperature
rises inversely, creating more uniform distributions. Otherwise, the
base temperature is maintained.

\subsection{3-Tier Organization}

The \texttt{EpistemicHead} sub-heads are organized in 3 tiers of
increasing complexity:

\begin{table}[H]
\centering
\caption{Sub-head organization by tiers.}
\label{tab:tiers}
\begin{tabular}{clrl}
\toprule
\textbf{Tier} & \textbf{Module} & \textbf{Params} & \textbf{Output} \\
\midrule
\multirow{5}{*}{\textbf{Core}} & Q1Gate & ~49K & $q_1$ [B,T,1] \\
& Q2Gate & ~49K & $q_2$ [B,T,1] \\
& DRM Manifold & ~16K & coords [B,T,5] \\
& MAD Confidence & ~2K & confidence [B,T,1] \\
& VI (Intentionality) & ~5K & $\phi$, $\bm{v}$ \\
\midrule
\multirow{3}{*}{\textbf{Tier 1}} & EidosDecay & ~368 & axis\_balance [B,T,5] \\
& Filosofia3 & ~373 & conflict [B,T,1] \\
& SelfModel & ~982 & mood, energy, drives \\
\midrule
\multirow{4}{*}{\textbf{Tier 2}} & TaskHead & ~50K & task\_probs [B,T,9] \\
& AmbiguityHead & ~50K & ambiguity [B,T,1] \\
& PlasticityGate & ~12K & plasticity [B,T,1] \\
& FrontierHead & ~129 & frontier [B,T,1] \\
\midrule
\multirow{3}{*}{\textbf{Tier 3}} & HumanStateHead & ~25K & psi [B,T,5] \\
& CausalState & ~26K & conditioning \\
& ContrastiveHead & ~200K & divergence [B,T,1] \\
\bottomrule
\end{tabular}
\end{table}

The total overhead of the sub-heads is {\raise.17ex\hbox{$\scriptstyle\sim$}}364K
parameters, representing \textbf{$<$0.3\%} of the 354M model. All
modules have \texttt{enable\_*} flags for conditional activation.

\subsection{EpistemicHead Forward}

The forward pass is executed sequentially within each tier, with
explicit dependencies:

\begin{algorithm}[H]
\caption{EpistemicHead Forward}
\label{alg:epistemic_forward}
\begin{algorithmic}[1]
\REQUIRE $\bh \in \R^{B \times T \times d}$, $\bm{A}$ (attention patterns)
\STATE \textbf{// Core}
\STATE $q_1 \leftarrow \text{Q1Gate}(\bh)$
\STATE $q_2 \leftarrow \text{Q2Gate}(\bh)$
\STATE $\bm{c} \leftarrow \text{ManifoldEmb}(\bh)$ \COMMENT{5D coords}
\STATE $\bG \leftarrow \text{MetricTensor}()$ \COMMENT{SPD metric tensor}
\STATE $d^2 \leftarrow \text{GeodesicDist}(\bm{c}, \bG)$ \COMMENT{distance}
\STATE $\kappa \leftarrow \text{MADConfidence}(\bm{c})$ \COMMENT{confidence}
\STATE $\phi \leftarrow \text{PhiField}(\bm{c}, \kappa)$ \COMMENT{health}
\STATE $\bm{v} \leftarrow \text{VI}(\phi, \kappa)$ \COMMENT{correction}
\STATE $\tau \leftarrow \text{AdaptiveTemp}(q_1, q_2)$ \COMMENT{temperature}
\STATE \textbf{// Tier 1} (if enabled)
\STATE $\bm{e} \leftarrow \text{EidosDecay}(\bm{c}, \kappa)$
\STATE $\gamma \leftarrow \text{Filosofia3}(\phi_{\text{comp}}, \kappa)$
\STATE $\bm{s} \leftarrow \text{SelfModel}(\bh, q_2, \phi, \kappa)$
\STATE \textbf{// Tiers 2--3} (omitted for brevity)
\RETURN $\mathcal{T}$ (full tomography)
\end{algorithmic}
\end{algorithm}
