% ============================================================
% Section 5: MAD - Metric-Aware Distance and Confidence
% ============================================================

\section{MAD: Confidence via Metric-Aware Distance}
\label{sec:mad}

The Metric-Aware Distance (MAD) system translates geodesic distances on
the 5D manifold into calibrated confidence, using a learned scale
parameter $\tau^2$ via backpropagation.

\subsection{Gaussian Confidence}

A token's confidence is defined as a Gaussian decay of the geodesic
distance to the manifold centroid:

\begin{equation}
    \kappa(\bm{c}) = \exp\left(-\frac{d_{\bG}^2(\bm{c}, \bm{\mu})}{2\tau^2}\right)
    \label{eq:mad_conf}
\end{equation}

where $d_{\bG}^2$ is the squared geodesic distance (\Cref{eq:geodesic}),
$\bm{\mu}$ is the \textbf{truth anchor} (fixed at
$[0.1, 0.1, 0.5, 0.9, 0.9]$, representing low uncertainty and high
response quality), and $\tau^2$ is the scale parameter.

The distance computation uses anisotropic (per-axis) $\tau^2$ by default:

\begin{equation}
    d_{\tau}^2(\bm{c}, \bm{\mu}) = \sum_{k=0}^{4} \frac{(c_k - \mu_k)^2}{\tau^2_k}
\end{equation}

where each axis $k$ has its own learnable scale $\tau^2_k$.

\subsection{Bayesian $\tau^2$}

The $\tau^2$ parameter is parametrized in log-scale to ensure positivity:

\begin{equation}
    \tau^2 = \exp(\log \tau^2_{\text{param}})
\end{equation}

where $\log \tau^2_{\text{param}}$ is a learned \texttt{nn.Parameter}
initialized at $0.0$, so $\tau^2 = \exp(0) = 1.0$ at the start of
training. This neutral initialization allows the model to freely
adjust the scale in either direction via backpropagation.

The parameter is clamped during forward to ensure
$\tau^2 \in [0.01, 10.0]$, preventing degenerate solutions.

\subsection{Calibration via Loss}

Calibration is enforced by the \texttt{MADCalibrationLoss}, which uses
Binary Cross-Entropy between confidence and accuracy:

\begin{equation}
    \calL_{\text{MAD}} = -\frac{1}{|\calS|}\sum_{t \in \calS} \Big[
        y_t \log \kappa_t + (1 - y_t) \log(1 - \kappa_t)
    \Big]
    \label{eq:mad_loss}
\end{equation}

where $y_t = \mathbb{1}[\hat{x}_t = x_t^*]$ is the correct prediction
indicator and $\calS$ is the set of valid tokens (after masking).

\begin{theorem}[Calibration at Equilibrium]
\label{thm:calibration}
At the minimum of \Cref{eq:mad_loss}, the confidence $\kappa$ is
calibrated:

\begin{equation}
    \E[\kappa_t \mid \kappa_t = p] = p, \quad \forall p \in [0, 1]
\end{equation}

i.e., tokens with confidence $p$ are correct with probability $p$.
\end{theorem}

\begin{proof}
BCE is minimized when $\kappa_t = P(y_t = 1 | \bm{c}_t)$ (well-known
property of binary cross-entropy). Therefore, at equilibrium, $\kappa_t$
estimates the true probability of correctness, which is exactly the
definition of calibration.
\end{proof}

\subsection{Gradient Flow}

The gradient of the MAD loss flows to the $\log \tau^2$ parameter:

\begin{equation}
    \frac{\partial \calL_{\text{MAD}}}{\partial \log \tau^2} =
    \frac{\partial \calL_{\text{MAD}}}{\partial \kappa} \cdot
    \frac{\partial \kappa}{\partial d^2} \cdot
    \frac{\partial d^2}{\partial \tau^2} \cdot
    \frac{\partial \tau^2}{\partial \log \tau^2}
\end{equation}

\begin{align}
    \frac{\partial \kappa}{\partial d^2} &= -\frac{\kappa}{2\tau^2} \\
    \frac{\partial d^2}{\partial \tau^2} &= 0 \quad \text{($d^2$ does not depend on $\tau^2$)} \\
    \frac{\partial \kappa}{\partial \tau^2} &= \frac{d^2 \kappa}{2\tau^4}
\end{align}

Additionally, the gradient also flows to the coordinates $\bm{c}$
(via $d^2$), to the metric tensor $\bG$ (via $d^2$), and to the
embedding MLP weights (\Cref{eq:manifold_emb}).

\subsection{Implementation with Autocast}

For numerical stability with mixed precision (bf16), the BCE is
implemented via \texttt{binary\_cross\_entropy\_with\_logits}:

\begin{equation}
    \text{logit}(\kappa) = \log\frac{\kappa}{1 - \kappa}
\end{equation}

followed by \texttt{F.binary\_cross\_entropy\_with\_logits(logit, target)},
which is numerically stable under autocast.
