% ============================================================
% Section 7: MPC Navigator
% ============================================================

\section{MPC Navigator on the Manifold}
\label{sec:mpc}

Model Predictive Control (MPC) extends the VI with trajectory planning
on the manifold, using beam search to find optimal sequences of
interventions.

\subsection{Problem Formulation}

The MPC navigator operates on the $\phi$ state
$\bm{s}_t = (\phi_1, \phi_2, \phi_3, \phi_4) \in \R^4$ and seeks the action sequence
$\bm{a}_{t:t+D}$ that minimizes a cumulative cost:

\begin{equation}
    \bm{a}^* = \arg\min_{\bm{a}_{t:t+D}} \sum_{k=0}^{D-1} C(\hat{\bm{s}}_{t+k}, a_{t+k})
    \label{eq:mpc_objective}
\end{equation}

where $D = 3$ is the planning horizon and $\hat{\bm{s}}$ are states
predicted by the transition model.

\subsection{Action Space}

The navigator has 12 intervention actions:

\begin{table}[H]
\centering
\caption{MPC intervention actions.}
\label{tab:mpc_actions}
\begin{tabular}{clcc}
\toprule
\textbf{ID} & \textbf{Action} & \textbf{Cost} & \textbf{Effect} \\
\midrule
0 & NOOP & 0.00 & Natural drift ($\times 0.995$) \\
1--5 & INJECT\_AXIS\_$k$ & 0.04 & Reinforce axis $k$ ($\phi_1$+0.05, $\phi_3$+0.03) \\
6 & INJECT\_WEAKEST & 0.06 & Reinforce weakest axis ($\phi_1$+0.07) \\
7 & INJECT\_TWO & 0.08 & Reinforce two weakest axes ($\phi_1$+0.10) \\
8 & CONF\_LIGHT & 0.03 & Light $\phi_4$ correction (5\%) \\
9 & CONF\_MEDIUM & 0.06 & Medium $\phi_4$ correction (10\%) \\
10 & CONF\_STRONG & 0.10 & Strong $\phi_4$ correction (20\%) \\
11 & RESET\_25 & 0.15 & 25\% reset toward mean \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Transition Model}

The transition model predicts the next state given an action:

\begin{equation}
    \hat{\bm{s}}_{t+1} = f_\theta(\bm{s}_t, a_t)
\end{equation}

implemented as a residual MLP that combines an analytical transition
with a learned correction:

\begin{equation}
    f_\theta(\bm{s}, a) = f_{\text{anal}}(\bm{s}, a) + \eta \cdot \text{MLP}_{16 \to 32 \to 32 \to 4}([\bm{s}; \text{onehot}(a)])
\end{equation}

where $f_{\text{anal}}$ applies deterministic effects
(\Cref{tab:mpc_actions}) and $\eta$ is a learnable residual scale
(initialized at 0.1). The MLP uses GELU activation with Tanh output,
taking $4 + 12 = 16$ input dimensions.

\subsection{Cost Function}

The cost at each step combines 4 terms:

The scoring depends on the operating mode:

\begin{equation}
    \text{score} = \begin{cases}
        \phi_{\text{final}} - 0.20 \cdot C_{\text{total}} - 0.02 \cdot n_{\text{changes}} & \text{(RECOVERY)} \\
        0.5 \cdot \max(0, \phi_{\text{final}} - 0.45) - C_{\text{total}} - 0.02 \cdot n_{\text{changes}} & \text{(MAINTENANCE)}
    \end{cases}
    \label{eq:mpc_cost}
\end{equation}

where $C_{\text{total}} = \sum_k c(a_k)$ is the cumulative action cost
(\Cref{tab:mpc_actions}) and $n_{\text{changes}}$ counts action transitions
(smoothness penalty).

\subsection{Beam Search}

The search space of $12^D = 1728$ trajectories is explored by beam
search with width $K = 4$:

\begin{algorithm}[H]
\caption{MPC Beam Search}
\label{alg:mpc_beam}
\begin{algorithmic}[1]
\REQUIRE Current state $\bm{s}_0$, model $f_\theta$, cost $C$
\STATE $\text{beams} \leftarrow \{(\bm{s}_0, [], 0)\}$ \COMMENT{(state, actions, cost)}
\FOR{$d = 0$ \textbf{to} $D-1$}
    \STATE $\text{candidates} \leftarrow \emptyset$
    \FOR{$(\bm{s}, \bm{a}, c)$ \textbf{in} beams}
        \FOR{$a \in \{0, \ldots, 11\}$}
            \STATE $\hat{\bm{s}} \leftarrow f_\theta(\bm{s}, a)$
            \STATE $\hat{c} \leftarrow c + C(\hat{\bm{s}}, a)$
            \STATE $\text{candidates} \leftarrow \text{candidates} \cup \{(\hat{\bm{s}}, \bm{a} \cup \{a\}, \hat{c})\}$
        \ENDFOR
    \ENDFOR
    \STATE $\text{beams} \leftarrow \text{top-}K(\text{candidates}, \text{by cost})$
\ENDFOR
\RETURN first action of the lowest-cost beam
\end{algorithmic}
\end{algorithm}

\subsection{$\phi$-Floor Constraint}

The navigator operates in two modes based on the $\phi$ level:

\begin{equation}
    \text{mode} = \begin{cases}
        \text{MAINTENANCE} & \text{if } \phi \geq \phi_{\text{floor}} = 0.45 \\
        \text{RECOVERY} & \text{if } \phi < \phi_{\text{floor}}
    \end{cases}
\end{equation}

In MAINTENANCE mode, action cost has higher weight (minimize
interventions). In RECOVERY mode, health cost dominates (maximize
$\phi$ quickly).

\subsection{Integration with VI}

The MPC operates on top of the VI, not as a replacement. The beam
search result generates a navigation hint that is combined with the
VI correction via weighted blending:

\begin{equation}
    \bm{v}_{\text{final}} = (1 - \lambda_{\text{nav}}) \bm{v}_{\text{VI}} + \lambda_{\text{nav}} \bm{v}_{\text{MPC}}
\end{equation}

with $\lambda_{\text{nav}} = 0.3$.
