% ============================================================
% Section 8: Extension Modules (9 modules)
% ============================================================

\section{Extension Modules}
\label{sec:extensions}

Beyond the epistemic core (DRM, MAD, VI, MPC), \aletheion{} includes
9 extension modules organized in 3 tiers. Each module is a trainable
\texttt{nn.Module} with an \texttt{enable\_*} flag for conditional
activation.

% -------------------------------------------------------
\subsection{Tier 1: High Impact, Low Cost ({\raise.17ex\hbox{$\scriptstyle\sim$}}1.7K params)}
% -------------------------------------------------------

\subsubsection{EidosDecay}
\label{sec:eidos}

EidosDecay monitors and corrects the balance of the 5 manifold axes.
Over-represented axes undergo decay (attenuation) while under-represented
axes receive reinforcement:

\begin{align}
    r_k &= \frac{\text{std}(c_k)}{\max_j \text{std}(c_j) + \epsilon} \label{eq:eidos_ratio} \\
    w_k^{\text{decay}} &= \sigma\big(\text{MLP}_{\text{decay}}(\bm{r})\big)_k \cdot \alpha_{\text{base}} \\
    w_k^{\text{reinforce}} &= \sigma\big(\text{MLP}_{\text{reinforce}}(\bm{r})\big)_k \cdot \beta_{\text{base}}
\end{align}

with $\alpha_{\text{base}} = 0.95$ and $\beta_{\text{base}} = 1.05$. The
result is an \texttt{axis\_balance} vector $\in \R^{B \times T \times 5}$
indicating the relative weight of each axis.

\paragraph{Dream Mode.} During ``dream'' phases (special training),
intensity is amplified by $\delta_{\text{dream}} = 3.0$:

\begin{equation}
    w_k^{\text{reinforce}} = \sigma(\cdot) \cdot \beta_{\text{base}} \cdot \delta_{\text{dream}}
\end{equation}

\paragraph{Parameters:} {\raise.17ex\hbox{$\scriptstyle\sim$}}368 (2 MLPs $5 \to 16 \to 5$ + 1 projection $5 \to 1$).

\subsubsection{Filosofia3: $\phi$-$\psi$ Conflict}
\label{sec:filosofia3}

The Filosofia3 module detects and quantifies conflicts between the
epistemic state ($\phi$) and the pragmatic state ($\psi$) of the model.

Conflict is measured via two pathways:

\paragraph{Analytical Pathway.}
Computes deltas between adjacent tokens and calculates cosine similarity:

\begin{align}
    \Delta\phi_t &= \phi_{t} - \phi_{t-1} \in \R^4 \\
    \Delta\psi_t &= (\kappa_t - \kappa_{t-1}) \cdot \bm{p} \in \R^4
\end{align}

where $\bm{p} = (0.1, 0.1, 0.1, 0.7)$ is the quality projection.
The analytical conflict is:

\begin{equation}
    \gamma_{\text{anal}} = \frac{1}{2}\left(1 - \frac{\Delta\phi \cdot \Delta\psi}{\|\Delta\phi\| \|\Delta\psi\| + \epsilon}\right) \in [0, 1]
\end{equation}

\paragraph{Learned Pathway.}
An MLP processes the concatenation $[\Delta\phi; \Delta\psi] \in \R^8$:

\begin{equation}
    \gamma_{\text{learn}} = \sigma\big(\text{MLP}_{8 \to 16 \to 1}([\Delta\phi; \Delta\psi])\big)
\end{equation}

\paragraph{Final Blend.}
\begin{equation}
    \gamma = 0.7 \cdot \gamma_{\text{anal}} + 0.3 \cdot \gamma_{\text{learn}}
\end{equation}

\paragraph{Operating Modes.}
A second MLP classifies the mode into 4 categories via softmax:

\begin{equation}
    \bm{m} = \text{softmax}\big(\text{MLP}_{8 \to 16 \to 4}([\Delta\phi; \Delta\psi])\big)
\end{equation}

\begin{enumerate}
    \item \textbf{ALIGNED} ($m_0$): $\phi$ and $\psi$ agree
    \item \textbf{CONFLICT\_TOLERATED} ($m_1$): Acceptable conflict
    \item \textbf{SIGNAL\_HUMAN} ($m_2$): Conflict requires signaling
    \item \textbf{RECOVERY} ($m_3$): Critical conflict
\end{enumerate}

\paragraph{Parameters:} {\raise.17ex\hbox{$\scriptstyle\sim$}}373 (2 MLPs + buffers).

\subsubsection{SelfModel: Cognitive Self-Model}
\label{sec:selfmodel}

The SelfModel maintains an internal model of the cognitive state,
producing 4 signals:

\begin{align}
    \text{mood} &= \tanh\big(\text{MLP}_{3 \to 32 \to 1}(\bar{\kappa}, \phi, p_{\text{success}})\big) \in [-1, 1] \\
    \text{curiosity} &= \sigma\big(\text{MLP}_{1 \to 8 \to 1}(q_2)\big) \in [0, 1] \\
    \text{energy} &= \sigma(\bm{w}^\top \bh) \cdot \text{decay}(t) \in [0, 1] \\
    \text{drives} &= \sigma\big(\text{MLP}_{3 \to 8 \to 3}(\text{curiosity}, \text{mood}, \text{energy})\big) \in [0, 1]^3
\end{align}

where $\text{decay}(t) = 1 - 0.3 \cdot t / T$ simulates energy depletion
along the sequence, and the 3 drives are: curiosity, mastery, and
efficiency.

\paragraph{Parameters:} {\raise.17ex\hbox{$\scriptstyle\sim$}}982 (dependent on $d_{\text{model}}$).

% -------------------------------------------------------
\subsection{Tier 2: Medium Impact, Medium Cost ({\raise.17ex\hbox{$\scriptstyle\sim$}}112K params)}
% -------------------------------------------------------

\subsubsection{Grounding: Task Classification and Ambiguity}
\label{sec:grounding}

\paragraph{TaskClassificationHead.}
Classifies each token into 9 task types via MLP:

\begin{equation}
    \bm{p}_{\text{task}} = \text{softmax}\big(\text{MLP}_{d \to 64 \to 9}(\bh)\big) \in \R^{B \times T \times 9}
\end{equation}

The 9 types are: FACTUAL, ANALYTICAL, CREATIVE, CONVERSATIONAL,
INSTRUCTIONAL, MATHEMATICAL, CODING, SUMMARIZATION, EPISTEMIC.

\paragraph{AmbiguityHead.}
Detects ambiguity at two levels:

\begin{align}
    \text{level} &= \sigma\big(\text{MLP}_{d \to 32 \to 1}(\bh)\big) \in [0, 1] \\
    \text{type} &= \text{softmax}\big(\text{MLP}_{d \to 32 \to 5}(\bh)\big) \in \R^5
\end{align}

The 5 ambiguity types: PRONOUN, SCOPE, LEXICAL, STRUCTURAL, REFERENTIAL.

\paragraph{Total parameters:} {\raise.17ex\hbox{$\scriptstyle\sim$}}99.4K.

\subsubsection{PlasticityGate}
\label{sec:plasticity}

Models the model's remaining plasticity along the sequence, simulating
cognitive resource depletion:

\begin{align}
    \text{cost}_t &= \text{Softplus}\big(\text{MLP}_{d \to 16 \to 1}(\bh_t)\big) \\
    \text{budget}_t &= 1.0 - 0.3 \cdot \frac{t}{T} \quad \text{(positional budget)} \\
    \text{plast}_t &= \text{clamp}\Big(\text{budget}_t - \delta \cdot \sum_{k=1}^{t} \text{cost}_k, \ 0, \ 1\Big) \\
    \text{gate}_t &= \sigma\big(\text{MLP}_{2 \to 8 \to 1}(\text{plast}_t, s_t)\big)
\end{align}

with $\delta = 0.02$ depletion rate.

\paragraph{Parameters:} {\raise.17ex\hbox{$\scriptstyle\sim$}}12.3K.

\subsubsection{MPL: Density and Frontier}
\label{sec:mpl}

\paragraph{DensityTracker.}
Maintains a sparse 5D hash grid that estimates visitation density in
each manifold region. Operates with \texttt{@torch.no\_grad}:

\begin{equation}
    \rho(\bm{c}) = \frac{1}{N} \sum_{i=1}^{N} K_h(\bm{c} - \bm{c}_i)
\end{equation}

with Gaussian kernel $K_h$ and bandwidth $h = 0.15$.

\paragraph{FrontierHead.}
Identifies unexplored manifold regions:

\begin{equation}
    f_{\text{frontier}} = \sigma\big(\text{MLP}_{6 \to 16 \to 1}([\bm{c}; \rho(\bm{c})])\big) \cdot (1 - \rho(\bm{c}))
\end{equation}

\paragraph{Parameters:} {\raise.17ex\hbox{$\scriptstyle\sim$}}129 (FrontierHead) + 0 (DensityTracker).

% -------------------------------------------------------
\subsection{Tier 3: Advanced Components ({\raise.17ex\hbox{$\scriptstyle\sim$}}250K params)}
% -------------------------------------------------------

\subsubsection{MOPsi: $\psi$ Orientation Module}
\label{sec:mopsi}

Models the ``human'' 5D state and mediates $\phi$-$\psi$ conflict:

\begin{align}
    \bpsi &= \sigma\big(\text{MLP}_{d \to 32 \to 5}(\bh)\big) \in [0, 1]^5 \\
    \psi_{\text{score}} &= \sigma\big(\text{MLP}_{10 \to 32 \to 1}([\bpsi; \bm{c}])\big) \in [0, 1]
\end{align}

The \texttt{PhiPsiMediator} resolves conflicts:

\begin{equation}
    \text{mediation} = \sigma\big(\text{MLP}_{3 \to 8 \to 1}(\phi, \psi_{\text{score}}, \gamma)\big)
\end{equation}

\paragraph{Parameters:} {\raise.17ex\hbox{$\scriptstyle\sim$}}25.3K + {\raise.17ex\hbox{$\scriptstyle\sim$}}41.

\subsubsection{CausalState: Causal Conditioning}
\label{sec:causal}

Allows conditioning generation on an external state vector
$\bm{z} \in \R^4$:

\begin{equation}
    \bh' = \bh + g \cdot \text{MLP}_{4 \to 32 \to d}(\bm{z})
\end{equation}

where $g = \sigma(\bm{w}^\top [\bh; \text{MLP}(\bm{z})])$ is a learned
gate. The \texttt{PolicyBinding} maps states to generation parameters
(temperature, max\_tokens).

\paragraph{Parameters:} {\raise.17ex\hbox{$\scriptstyle\sim$}}25.5K (StateConditioning) + 0 (PolicyBinding).

\subsubsection{Metacognitive: Contrastive Evaluation}
\label{sec:metacognitive}

The ContrastiveHead projects hidden states into two ``views'' and
measures divergence:

\begin{align}
    \bm{v}_a &= \text{MLP}_{d \to 128 \to d/2}(\bh) \\
    \bm{v}_b &= \text{MLP}_{d \to 128 \to d/2}(\bh) \\
    \text{div} &= \sigma\big(\text{MLP}_{d \to 32 \to 1}([\bm{v}_a; \bm{v}_b])\big) \in [0, 1]
\end{align}

High divergence indicates the model has ``two opinions'' about the
token, signaling metacognitive uncertainty.

\paragraph{Parameters:} {\raise.17ex\hbox{$\scriptstyle\sim$}}200K.

\subsection{Total Overhead}

\begin{table}[H]
\centering
\caption{Extension module parameters.}
\label{tab:overhead}
\begin{tabular}{lrl}
\toprule
\textbf{Module} & \textbf{Params} & \textbf{\% of Model} \\
\midrule
EidosDecay & 368 & $<$0.001\% \\
Filosofia3 & 373 & $<$0.001\% \\
SelfModel & 982 & $<$0.001\% \\
Grounding (2 heads) & 99.4K & 0.028\% \\
PlasticityGate & 12.3K & 0.003\% \\
MPL (Frontier) & 129 & $<$0.001\% \\
MOPsi & 25.3K & 0.007\% \\
CausalState & 25.5K & 0.007\% \\
Metacognitive & 200K & 0.056\% \\
\midrule
\textbf{Total} & \textbf{{\raise.17ex\hbox{$\scriptstyle\sim$}}364K} & \textbf{$<$0.3\%} \\
\bottomrule
\end{tabular}
\end{table}

All modules are optional and can be enabled/disabled via configuration
without altering the base transformer architecture.
