% ============================================================
% Section 10: Continual Learning
% ============================================================

\section{Continual Learning}
\label{sec:continual}

\aletheion{} integrates two complementary Continual Learning (CL)
mechanisms to preserve knowledge across multiple training phases:
Elastic Weight Consolidation (EWC) and Experience Replay.

\subsection{Motivation}

Catastrophic forgetting \cite{mccloskey1989catastrophic} is the
phenomenon where neural networks lose knowledge of previous tasks
when trained on new ones. In an LLM with an epistemic system, this is
particularly problematic:

\begin{itemize}
    \item The 5D manifold may ``forget'' its calibrated geometry
    \item The $\tau^2$ parameter may become decalibrated
    \item Epistemic sub-heads may lose learned signals
\end{itemize}

\subsection{EWC: Elastic Weight Consolidation}
\label{sec:ewc}

EWC \cite{kirkpatrick2017ewc} protects important parameters by adding
a quadratic penalty weighted by the diagonal Fisher Information Matrix
(FIM).

\subsubsection{Fisher Information Matrix}

After each training phase, we estimate the diagonal FIM:

\begin{equation}
    F_i = \E\left[\left(\frac{\partial \calL}{\partial \theta_i}\right)^2\right]
    \approx \frac{1}{N} \sum_{n=1}^{N} \left(\frac{\partial \calL(\bx^{(n)})}{\partial \theta_i}\right)^2
    \label{eq:fisher}
\end{equation}

where $N = 256$ samples from the dataset (configurable via
\texttt{ewc\_fisher\_samples}).

$F_i$ measures the \textbf{importance} of parameter $\theta_i$: high
values indicate the parameter is critical for the current task.

\subsubsection{EWC Penalty}

In the subsequent phase, the total loss includes:

\begin{equation}
    \calL_{\text{EWC}} = \frac{\lambda_{\text{EWC}}}{2} \sum_{i} F_i \cdot (\theta_i - \theta_i^*)^2
    \label{eq:ewc_loss}
\end{equation}

where $\theta_i^*$ are the parameters after the previous phase and
$\lambda_{\text{EWC}} = 100.0$ controls the rigidity.

\subsubsection{Online EWC}

For multiple phases, we use online EWC \cite{schwarz2018progress}
with exponential accumulation:

\begin{align}
    \bar{F}_i^{(k)} &= \gamma \cdot \bar{F}_i^{(k-1)} + F_i^{(k)} \label{eq:ewc_online} \\
    \theta_i^{*(k)} &= \theta_i^{(k)}
\end{align}

with $\gamma = 0.9$ (decay). This avoids the linear growth of terms
that occurs with naive multi-task EWC.

\subsubsection{Trainer Integration}

\Cref{eq:ewc_loss} is added to the total loss after the composite loss
computation:

\begin{equation}
    \calL_{\text{final}} = \calL_{\text{total}} + \calL_{\text{EWC}}
\end{equation}

The \texttt{consolidate\_phase()} method must be called after each phase:

\begin{lstlisting}[caption={EWC usage in training.}]
# Phase 1: Base training
trainer.train()
trainer.consolidate_phase()  # Compute Fisher

# Phase 2: New dataset (EWC protects phase 1)
trainer.train_loader = new_loader
trainer.train()
trainer.consolidate_phase()  # Accumulate Fisher (online)

# Phase 3: More data (EWC protects phases 1+2)
trainer.train_loader = another_loader
trainer.train()
\end{lstlisting}

\subsection{Experience Replay}
\label{sec:replay}

Experience Replay complements EWC by maintaining a buffer of previous
samples that are mixed into the current batch.

\subsubsection{Reservoir Sampling}

The buffer uses Vitter's reservoir sampling \cite{vitter1985reservoir}
to maintain a uniform sample of fixed size:

\begin{algorithm}[H]
\caption{Reservoir Sampling}
\begin{algorithmic}[1]
\REQUIRE Buffer $\calB$ of capacity $M$, sample $\bx_n$ ($n$-th)
\IF{$|\calB| < M$}
    \STATE $\calB \leftarrow \calB \cup \{\bx_n\}$
\ELSE
    \STATE $j \sim \text{Uniform}(0, n)$
    \IF{$j < M$}
        \STATE $\calB[j] \leftarrow \bx_n$
    \ENDIF
\ENDIF
\end{algorithmic}
\end{algorithm}

This ensures that each sample seen up to time $n$ has probability
$M/n$ of being in the buffer, without needing to store the entire
history.

\subsubsection{Mix Batch}

At each training step, a fraction $\rho = 0.1$ of the batch is replaced
with buffer samples:

\begin{equation}
    \tilde{\calB}_t = (1 - \rho) \cdot \calB_t^{\text{new}} \cup \rho \cdot \text{sample}(\calB_{\text{replay}})
\end{equation}

\subsubsection{CPU Storage}

The buffer is stored in CPU memory to avoid consuming VRAM:

\begin{equation}
    \text{mem}_{\text{buffer}} = M \cdot T \cdot 2 \text{ bytes} \approx 10000 \cdot 1024 \cdot 2 \approx 20 \text{ MB}
\end{equation}

Samples are transferred to GPU only when selected for the mix batch.

\subsection{EWC + Replay Synergy}

The two mechanisms are complementary:

\begin{itemize}
    \item \textbf{EWC} protects in \textit{parameter space}: prevents
    important weights from changing
    \item \textbf{Replay} protects in \textit{data space}: keeps
    examples from the past in training
\end{itemize}

Together, they reduce forgetting more than either one alone
\cite{chaudhry2019continual}.

\subsection{Checkpointing with CL State}

The CL state is saved alongside the model checkpoint:

\begin{lstlisting}[caption={Checkpoint contents.}]
checkpoint = {
    'model': model.state_dict(),
    'optimizer': optimizer.state_dict(),
    'scheduler': scheduler.state_dict(),
    'global_step': step,
    'ewc_state': ewc.state_dict_ewc(),   # Fisher + ref params
    'replay_state': replay.state_dict(),  # Buffer + counters
}
\end{lstlisting}

This allows resuming training with CL state intact.

\subsection{Configuration}

\begin{table}[H]
\centering
\caption{Continual Learning parameters.}
\label{tab:cl_config}
\begin{tabular}{lrl}
\toprule
\textbf{Parameter} & \textbf{Default} & \textbf{Description} \\
\midrule
\texttt{enable\_ewc} & true & Enable EWC \\
\texttt{ewc\_lambda} & 100.0 & EWC penalty weight \\
\texttt{ewc\_online} & true & Online Fisher accumulation \\
\texttt{ewc\_gamma} & 0.9 & Accumulated Fisher decay \\
\texttt{ewc\_fisher\_samples} & 256 & Samples for Fisher estimation \\
\texttt{enable\_replay} & true & Enable Experience Replay \\
\texttt{replay\_buffer\_size} & 10000 & Buffer capacity \\
\texttt{replay\_mix\_ratio} & 0.1 & Replay fraction in batch \\
\bottomrule
\end{tabular}
\end{table}
